---
title: "Convergence and diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastclogit)
set.seed(1)
```

At a few thousand rows, a conditional logit either converges or it does not,
and you rarely think about it. At tens of millions of rows with a sampling
offset and rare cells, "converged" stops being a yes or no and becomes a
question about *how*. This vignette is about answering that question.

It exists because getting it wrong has been expensive here. Two defects in this
package's line search went unnoticed for months: one let the optimiser accept a
wildly oversized first step and wander into a region where the softmax
saturates and the gradient freezes; the other accepted steps that improved
nothing, so a fit walked downhill for ninety iterations at seventy seconds
each. Both produced output that looked like a fit.

## The three things every fit reports

```{r}
sim <- simulate_clogit_data(n_egos = 2000, n_alts = 20, seed = 42)
fit <- fastclogit(sim$X, sim$choice, sim$strata, offset = sim$offset)

fit$convergence_criterion
fit$convergence_message
```

And the full trace, one row per Newton iteration:

```{r}
fit$iter_log
```

`step_size` and `halving_count` describe the step that *produced* the beta
evaluated on that row, so they are `NA` on iteration 1. `tier_fired` records
which convergence test passed, if any.

## The six routes, and what each one means

| `convergence_criterion` | Converged? | What it means |
|---|---|---|
| `primary` | yes | `max\|gradient\| < tol`. Unambiguous. |
| `secondary` | yes | Relative log-likelihood change, gradient, and the unhalved Newton step are all tight together. |
| `plateau` | yes, conditionally | The log-likelihood is flat over several iterations, the gradient is below `tier3_grad_floor`, and the optimiser shows evidence of struggling. |
| `flat_optimum` | yes | No step improves the log-likelihood and the gradient is already small. That is what being at a maximum means. |
| `iter_max` | **no** | Hit `max_iter`. |
| `line_search` | **no** | No step improves the log-likelihood *and* the gradient is still large. The Newton direction has stopped being an ascent direction. |

### Why `plateau` needs side-conditions

`plateau` matches `survival::clogit`'s relative-log-likelihood criterion
(`1e-9`). On its own that is too permissive: a badly stuck optimiser also stops
moving. So tier-3 additionally requires the change to persist over
`tier3_plateau_iters` consecutive iterations, the gradient to be below
`tier3_grad_floor`, and the previous iteration to show either step-halving or a
very small step.

That last condition matters more than it looks. It is evidence the optimiser is
on a genuinely flat ridge rather than simply finished, and a bug in the line
search can *manufacture* it. Treat a `plateau` fit as a fit worth a second
look, not as a failure.

### `flat_optimum` versus `line_search`

These two look identical from inside the loop: no step improves the
log-likelihood. The gradient distinguishes them, and getting that wrong was a
real incident. An early version of the guard treated every such case as a
failure, which relabelled finished fits as `FAILED` and silently skipped a
men's KHB decomposition. The rule now splits on `tier3_grad_floor`, the same
threshold the plateau rule uses, so the two agree by construction.

## Reading a trace that went wrong

Four signatures are worth recognising.

**The log-likelihood jumps by orders of magnitude between iterations 1 and 2.**
The opening step was far too long. With the line search running on every
iteration this should no longer happen, but if you see it, stop and look at
your offset's within-stratum spread.

**`grad_max` is frozen at the same large value for many iterations.** The
softmax has saturated; the observed information has collapsed toward zero.
Nothing after that point recovers, and the fit should be discarded rather than
given more iterations.

**`halving_count` sits at 20 every iteration.** Either the direction is bad, or
the acceptance test is comparing rounding noise, which happens once
`|loglik|` is large enough that `1e-10` falls below one unit in the last place.
The kernel now scales that tolerance with `|loglik|`.

**The log-likelihood decreases slowly and monotonically.** This is the second
historic defect. It cannot happen now: a step that does not improve is refused.
`inst/validation` asserts this directly.

## "Did not converge" at a tiny gradient

A fit can hit `max_iter` with a gradient far below anything you would call
unconverged:

```{r}
small <- simulate_clogit_data(n_egos = 1000, n_alts = 10,
                              use_offset = FALSE, seed = 7)
tight <- fastclogit(small$X, small$choice, small$strata,
                    tol = 1e-14, max_iter = 60L)
tight$convergence_criterion
max(abs(tight$gradient))
```

`tol` applies to `max|gradient|`, which is a sum over strata, so how small it
can be *computed* to be depends on the problem. Ask for 1e-14 and the optimiser
will keep trying to reach it long after it has arrived. The warning says so
explicitly. Relax `tol`; do not raise `max_iter`.

## The knobs

You will rarely touch these, and the defaults are what the papers run.

| Argument | Default | Change it when |
|---|---|---|
| `tol` | `1e-6` | The fit reports `iter_max` at a gradient you consider small. Relax it. |
| `max_iter` | `25` | A genuinely hard problem is still improving at the cap. Check `iter_log` first: if the log-likelihood is flat, more iterations will not help. |
| `tier3_enable` | `TRUE` | You want strictly gradient-based convergence, e.g. to reproduce pre-2026-06-17 behaviour. |
| `tier3_plateau_tol` | `1e-9` | Almost never. It matches `survival`'s `eps`; the extra strictness comes from the side-conditions, not from tightening this. |
| `tier3_plateau_iters` | `3` | You want more consecutive flat iterations before accepting a plateau. |
| `tier3_grad_floor` | `1e-2` | You want to refuse plateau convergence unless the gradient is smaller. Also governs `flat_optimum`. |
| `tier3_halving_floor`, `tier3_step_floor` | `2`, `1e-3` | Almost never. They define "the optimiser is struggling". |

## A checklist for a production fit

1. `fit$convergence_criterion` is `primary` or `flat_optimum` → done.
2. `secondary` → check `max(abs(fit$gradient))` is near `tol`.
3. `plateau` → look at `fit$iter_log`. Is the log-likelihood genuinely flat
   over the last few iterations, and is the gradient small in absolute terms?
4. `iter_max` → is the gradient already tiny? Then relax `tol`. Otherwise look
   for a saturated softmax.
5. `line_search` → do not use this fit.
6. Check `fit$vcov_singular`. If `TRUE`, some SEs rest on a ridge rather than
   on the data. Note that it does **not** fire for stratum-constant columns,
   which are a separate problem, see `?fastclogit`.
