---
title: "Getting Started with fastclogit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fastclogit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastclogit)
```

## Overview

`fastclogit` estimates conditional logit (McFadden's choice) models using a
memory-efficient Rcpp implementation. It is designed as a drop-in replacement
for `survival::clogit()` when your data is too large to fit in memory with
the standard approach.

The key difference: `clogit()` builds the full design matrix via
`model.matrix()`, which duplicates your data in memory. For a 90-million-row
dataset, this can require 80--120 GB of RAM. `fastclogit` builds the design
matrix column-by-column and passes it directly to C++, using roughly 5--10x
less memory.

## Simulating data

The package includes a data generator that creates synthetic discrete choice
data with a structure typical of large discrete-choice problems:

```{r simulate}
sim <- simulate_clogit_data(
  n_egos  = 1000,   # number of choice sets
  n_alts  = 50,     # alternatives per choice set
  use_offset = TRUE, # McFadden/Manski sampling correction
  seed    = 42
)

str(sim$data[1:5, ])
```

Each ego (chooser) has `n_alts` alternatives. Exactly one alternative per
choice set has `choice = 1`. The `correction` column contains the
McFadden/Manski offset that adjusts for stratified sampling.

## Fitting a model

### Formula interface (recommended)

```{r fclogit}
fit <- fclogit(
  choice ~ x1 + x2 + x3 + x4,
  data    = sim$data,
  strata  = "strata_id",
  cluster = "cluster_id",
  offset  = "correction"
)

summary(fit)
```

The formula works just like in `lm()` or `glm()`. Factor variables are
automatically dummy-coded (dropping the first level as reference).
Interactions are supported with the `:` operator.

### Matrix interface

If you prefer to construct the design matrix yourself:

```{r matrix-interface}
fit2 <- fastclogit(
  X       = sim$X,
  choice  = sim$choice,
  strata  = sim$strata,
  offset  = sim$offset,
  cluster = sim$cluster
)

# Same coefficients
all.equal(coef(fit2), sim$beta_true, tolerance = 0.3)
```

### Sparse-X path (factor-heavy designs at scale)

When the design matrix has many factor dummies and few non-zeros per row,
passing a `Matrix::sparseMatrix` triggers a CSR-walking Newton kernel
that avoids ever materialising the dense matrix. Coefficients agree with
the dense fit at machine epsilon; memory and runtime drop substantially.

```{r sparse, eval = requireNamespace("Matrix", quietly = TRUE)}
library(Matrix)

# Convert the same X to sparse and re-fit
X_sparse <- as(sim$X, "CsparseMatrix")
fit_sp <- fastclogit(
  X = X_sparse, choice = sim$choice, strata = sim$strata,
  offset = sim$offset, cluster = sim$cluster
)

# Bit-identical coefficients (tolerance is machine epsilon)
max(abs(coef(fit_sp) - coef(fit2)))
```

For very large factor models, an interactions-heavy design at 1M+ rows with
90+ columns at ~5% density, say, build the sparse design matrix directly from
a formula via `Matrix::sparse.model.matrix()` so the dense version is never
built:

```{r sparse-formula, eval = FALSE}
library(Matrix)
X_sparse <- sparse.model.matrix(
  ~ group * decade + age * decade + edu * decade + dist * decade,
  data = dt)[, -1, drop = FALSE]   # drop the intercept

fit <- fastclogit(X_sparse,
                  choice  = dt$chosen,
                  strata  = dt$choice_set_id,
                  cluster = dt$person_id)
```

See `vignette("large-scale")` for measured timings and memory across a range
of problem sizes, and for how to size a job before you queue it.

## Checking how the fit converged

Every fit reports the route it took, not just whether it finished:

```{r}
fit$convergence_criterion
fit$convergence_message
head(fit$iter_log)
```

`primary` and `flat_optimum` are unambiguous successes; `iter_max` and
`line_search` are failures. At small scale you will almost always see
`primary`. At tens of millions of rows the distinction matters a great deal,
and `vignette("convergence-and-diagnostics")` is about exactly that.

## Extracting results

Standard S3 methods work as expected:

```{r methods}
coef(fit)
confint(fit)
tidy_fastclogit(fit)  # broom-style data.frame
```

By default, `summary()`, `confint()`, and `tidy_fastclogit()` use clustered
robust standard errors when available. Pass `robust = FALSE` to use
model-based SEs instead.

## Also included

`khb_decompose()` implements the Kohler, Karlson & Holm (2011) decomposition of
total effects into direct and indirect parts for nested nonlinear models, built
on `fclogit()`. It is a specialised tool rather than part of the core workflow;
see `?khb_decompose` for a worked example.

## Comparison with survival::clogit

If you have `survival` installed, you can verify that `fastclogit` produces
the same estimates:

```{r compare, eval = requireNamespace("survival", quietly = TRUE)}
library(survival)
sim_small <- simulate_clogit_data(n_egos = 200, n_alts = 20, seed = 1)

# fastclogit
fit_fast <- fastclogit(sim_small$X, sim_small$choice, sim_small$strata,
                        offset = sim_small$offset, cluster = sim_small$cluster)

# survival::clogit
d <- sim_small$data
d$strata_id <- factor(d$strata_id)
d$cluster_id <- factor(d$cluster_id)
fit_clogit <- clogit(
  choice ~ x1 + x2 + x3 + x4 +
    f1_b + f1_c + f1_d +
    f2_b + f2_c + f2_d + f2_e +
    offset(correction) + strata(strata_id) + cluster(cluster_id),
  data = d,
  # clogit() defaults to method = "exact", which it refuses to combine with a
  # robust variance. "efron" is exact for 1:m matched data (one chosen
  # alternative per stratum), which is what conditional logit always is here.
  method = "efron"
)

# Compare coefficients
data.frame(
  term = names(coef(fit_fast)),
  fastclogit = round(coef(fit_fast), 6),
  clogit = round(coef(fit_clogit), 6),
  diff = round(coef(fit_fast) - coef(fit_clogit), 10)
)
```
