---
title: "Getting Started with fastclogit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fastclogit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastclogit)
```

## Overview

`fastclogit` estimates conditional logit (McFadden's choice) models using a
memory-efficient Rcpp implementation. It is designed as a drop-in replacement
for `survival::clogit()` when your data is too large to fit in memory with
the standard approach.

The key difference: `clogit()` builds the full design matrix via
`model.matrix()`, which duplicates your data in memory. For a 90-million-row
dataset, this can require 80--120 GB of RAM. `fastclogit` builds the design
matrix column-by-column and passes it directly to C++, using roughly 5--10x
less memory.

## Simulating data

The package includes a data generator that creates synthetic discrete choice
data with the structure commonly used in partner choice research:

```{r simulate}
sim <- simulate_clogit_data(
  n_egos  = 1000,   # number of choice sets
  n_alts  = 50,     # alternatives per choice set
  use_offset = TRUE, # McFadden/Manski sampling correction
  seed    = 42
)

str(sim$data[1:5, ])
```

Each ego (chooser) has `n_alts` alternatives. Exactly one alternative per
choice set has `choice = 1`. The `correction` column contains the
McFadden/Manski offset that adjusts for stratified sampling.

## Fitting a model

### Formula interface (recommended)

```{r fclogit}
fit <- fclogit(
  choice ~ lnDist + n_years_same_cfar + n_years_same_peorg + n_years_same_uni,
  data    = sim$data,
  strata  = "strata_id",
  cluster = "cluster_id",
  offset  = "correction"
)

summary(fit)
```

The formula works just like in `lm()` or `glm()`. Factor variables are
automatically dummy-coded (dropping the first level as reference).
Interactions are supported with the `:` operator.

### Matrix interface

If you prefer to construct the design matrix yourself:

```{r matrix-interface}
fit2 <- fastclogit(
  X       = sim$X,
  choice  = sim$choice,
  strata  = sim$strata,
  offset  = sim$offset,
  cluster = sim$cluster
)

# Same coefficients
all.equal(coef(fit2), sim$beta_true, tolerance = 0.3)
```

## Extracting results

Standard S3 methods work as expected:

```{r methods}
coef(fit)
confint(fit)
tidy_fastclogit(fit)  # broom-style data.frame
```

By default, `summary()`, `confint()`, and `tidy_fastclogit()` use clustered
robust standard errors when available. Pass `robust = FALSE` to use
model-based SEs instead.

## KHB mediation decomposition

The KHB method (Kohler, Karlson & Holm 2011) decomposes total effects into
direct and indirect (mediated) effects in nonlinear models. This is useful
when you want to know how much of variable X's effect operates through
mediator Z.

```{r khb}
# Create data with a known mediation structure
set.seed(123)
n_egos <- 500; n_alts <- 30; n <- n_egos * n_alts

d <- data.frame(
  strata_id = rep(1:n_egos, each = n_alts),
  x = rnorm(n),
  c1 = rnorm(n)
)
d$z <- 0.6 * d$x + rnorm(n)  # z is mediated by x (gamma = 0.6)
eta <- 0.5 * d$x + 0.8 * d$z + 0.3 * d$c1

# Generate choices
d$choice <- 0L
for (i in 1:n_egos) {
  rows <- ((i-1)*n_alts + 1):(i*n_alts)
  probs <- exp(eta[rows] - max(eta[rows]))
  probs <- probs / sum(probs)
  d$choice[rows[sample(n_alts, 1, prob = probs)]] <- 1L
}

result <- khb_decompose(
  data     = d,
  key_vars = "x",
  z_vars   = "z",
  controls = "c1",
  strata   = "strata_id",
  choice   = "choice",
  verbose  = FALSE
)

result$decomposition
```

The indirect effect captures the portion of x's effect that operates through
z. The confounding percentage tells you what fraction of x's total effect is
mediated.

## Comparison with survival::clogit

If you have `survival` installed, you can verify that `fastclogit` produces
the same estimates:

```{r compare, eval = requireNamespace("survival", quietly = TRUE)}
library(survival)
sim_small <- simulate_clogit_data(n_egos = 200, n_alts = 20, seed = 1)

# fastclogit
fit_fast <- fastclogit(sim_small$X, sim_small$choice, sim_small$strata,
                        offset = sim_small$offset, cluster = sim_small$cluster)

# survival::clogit
d <- sim_small$data
d$strata_id <- factor(d$strata_id)
d$cluster_id <- factor(d$cluster_id)
fit_clogit <- clogit(
  choice ~ lnDist + n_years_same_cfar + n_years_same_peorg + n_years_same_uni +
    Edudiff3_BothHigh + Edudiff3_BothLow + Edudiff3_Missing +
    AgeDiffcat_2 + AgeDiffcat_3 + AgeDiffcat_4 + AgeDiffcat_5 +
    offset(correction) + strata(strata_id) + cluster(cluster_id),
  data = d
)

# Compare coefficients
data.frame(
  term = names(coef(fit_fast)),
  fastclogit = round(coef(fit_fast), 6),
  clogit = round(coef(fit_clogit), 6),
  diff = round(coef(fit_fast) - coef(fit_clogit), 10)
)
```
