---
title: "Getting Started with fastclogit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fastclogit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastclogit)
```

## Overview

`fastclogit` estimates conditional logit (McFadden's choice) models using a
memory-efficient Rcpp implementation. It is designed as a drop-in replacement
for `survival::clogit()` when your data is too large to fit in memory with
the standard approach.

The key difference: `clogit()` builds the full design matrix via
`model.matrix()`, which duplicates your data in memory. For a 90-million-row
dataset, this can require 80--120 GB of RAM. `fastclogit` builds the design
matrix column-by-column and passes it directly to C++, using roughly 5--10x
less memory.

## Simulating data

The package includes a data generator that creates synthetic discrete choice
data with the structure commonly used in partner choice research:

```{r simulate}
sim <- simulate_clogit_data(
  n_egos  = 1000,   # number of choice sets
  n_alts  = 50,     # alternatives per choice set
  use_offset = TRUE, # McFadden/Manski sampling correction
  seed    = 42
)

str(sim$data[1:5, ])
```

Each ego (chooser) has `n_alts` alternatives. Exactly one alternative per
choice set has `choice = 1`. The `correction` column contains the
McFadden/Manski offset that adjusts for stratified sampling.

## Fitting a model

### Formula interface (recommended)

```{r fclogit}
fit <- fclogit(
  choice ~ lnDist + n_years_same_cfar + n_years_same_peorg + n_years_same_uni,
  data    = sim$data,
  strata  = "strata_id",
  cluster = "cluster_id",
  offset  = "correction"
)

summary(fit)
```

The formula works just like in `lm()` or `glm()`. Factor variables are
automatically dummy-coded (dropping the first level as reference).
Interactions are supported with the `:` operator.

### Matrix interface

If you prefer to construct the design matrix yourself:

```{r matrix-interface}
fit2 <- fastclogit(
  X       = sim$X,
  choice  = sim$choice,
  strata  = sim$strata,
  offset  = sim$offset,
  cluster = sim$cluster
)

# Same coefficients
all.equal(coef(fit2), sim$beta_true, tolerance = 0.3)
```

### Sparse-X path (factor-heavy designs at scale)

When the design matrix has many factor dummies and few non-zeros per row,
passing a `Matrix::sparseMatrix` triggers a CSR-walking Newton kernel
that avoids ever materialising the dense matrix. Coefficients agree with
the dense fit at machine epsilon; memory and runtime drop substantially.

```{r sparse, eval = requireNamespace("Matrix", quietly = TRUE)}
library(Matrix)

# Convert the same X to sparse and re-fit
X_sparse <- as(sim$X, "CsparseMatrix")
fit_sp <- fastclogit(
  X = X_sparse, choice = sim$choice, strata = sim$strata,
  offset = sim$offset, cluster = sim$cluster
)

# Bit-identical coefficients (tolerance is machine epsilon)
max(abs(coef(fit_sp) - coef(fit2)))
```

For very large factor models — for example, an interactions-heavy Paper-3
style design at 1M+ rows with 90+ columns at ~5% density — build the
sparse design matrix directly from a formula via
`Matrix::sparse.model.matrix()` so the dense version is never built:

```{r sparse-formula, eval = FALSE}
library(Matrix)
X_sparse <- sparse.model.matrix(
  ~ pair_gen_meso * decade + age * decade + edu * decade + lnDist * decade,
  data = dt)[, -1, drop = FALSE]   # drop the intercept

fit <- fastclogit(X_sparse,
                  choice  = dt$actualpartner,
                  strata  = dt$CoupleId,
                  cluster = dt$LopNrEgo)
```

In production benchmarks at 50,000 strata × 30 alters × 92 cols (~5%
density), the sparse path is 20× faster than the dense path (1.4 s vs
27.5 s) and uses 15% less memory. The gap widens with problem size:
at the Paper-3 Step 4 target scale, the sparse path takes the fit from
~95 minutes / ~225 GB to ~80 seconds / ~30 GB.

## Extracting results

Standard S3 methods work as expected:

```{r methods}
coef(fit)
confint(fit)
tidy_fastclogit(fit)  # broom-style data.frame
```

By default, `summary()`, `confint()`, and `tidy_fastclogit()` use clustered
robust standard errors when available. Pass `robust = FALSE` to use
model-based SEs instead.

## KHB mediation decomposition

The KHB method (Kohler, Karlson & Holm 2011) decomposes total effects into
direct and indirect (mediated) effects in nonlinear models. This is useful
when you want to know how much of variable X's effect operates through
mediator Z.

```{r khb}
# Create data with a known mediation structure
set.seed(123)
n_egos <- 500; n_alts <- 30; n <- n_egos * n_alts

d <- data.frame(
  strata_id = rep(1:n_egos, each = n_alts),
  x = rnorm(n),
  c1 = rnorm(n)
)
d$z <- 0.6 * d$x + rnorm(n)  # z is mediated by x (gamma = 0.6)
eta <- 0.5 * d$x + 0.8 * d$z + 0.3 * d$c1

# Generate choices
d$choice <- 0L
for (i in 1:n_egos) {
  rows <- ((i-1)*n_alts + 1):(i*n_alts)
  probs <- exp(eta[rows] - max(eta[rows]))
  probs <- probs / sum(probs)
  d$choice[rows[sample(n_alts, 1, prob = probs)]] <- 1L
}

result <- khb_decompose(
  data     = d,
  key_vars = "x",
  z_vars   = "z",
  controls = "c1",
  strata   = "strata_id",
  choice   = "choice",
  verbose  = FALSE
)

result$decomposition
```

The indirect effect captures the portion of x's effect that operates through
z. The confounding percentage tells you what fraction of x's total effect is
mediated.

## Comparison with survival::clogit

If you have `survival` installed, you can verify that `fastclogit` produces
the same estimates:

```{r compare, eval = requireNamespace("survival", quietly = TRUE)}
library(survival)
sim_small <- simulate_clogit_data(n_egos = 200, n_alts = 20, seed = 1)

# fastclogit
fit_fast <- fastclogit(sim_small$X, sim_small$choice, sim_small$strata,
                        offset = sim_small$offset, cluster = sim_small$cluster)

# survival::clogit
d <- sim_small$data
d$strata_id <- factor(d$strata_id)
d$cluster_id <- factor(d$cluster_id)
fit_clogit <- clogit(
  choice ~ lnDist + n_years_same_cfar + n_years_same_peorg + n_years_same_uni +
    Edudiff3_BothHigh + Edudiff3_BothLow + Edudiff3_Missing +
    AgeDiffcat_2 + AgeDiffcat_3 + AgeDiffcat_4 + AgeDiffcat_5 +
    offset(correction) + strata(strata_id) + cluster(cluster_id),
  data = d
)

# Compare coefficients
data.frame(
  term = names(coef(fit_fast)),
  fastclogit = round(coef(fit_fast), 6),
  clogit = round(coef(fit_clogit), 6),
  diff = round(coef(fit_fast) - coef(fit_clogit), 10)
)
```
