---
title: "Large-scale designs and restricted servers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Large-scale designs and restricted servers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastclogit)
```

How to run this package where `survival::clogit()` cannot go: factor-heavy
designs at tens of millions of rows, and servers where you cannot install
packages.

## When the sparse path pays

The sparse kernel builds a row-major (CSR) index of the design matrix once and
walks rows per stratum. It wins when the design has many factor dummies and few
non-zeros per row, which is the usual shape for interacted categorical models.
It does not win on dense continuous predictors, where it adds the CSR build for
no compression benefit.

Measured on one machine, one process per configuration, peak RSS from
`/usr/bin/time -l`:

| rows | cols | dense | sparse |
|---|---|---|---|
| 100,000 | 20 | 0.33 s / 0.30 GB | 0.05 s / 0.25 GB |
| 500,000 | 50 | 4.32 s / 1.24 GB | 0.26 s / 0.41 GB |
| 1,000,000 | 90 | 21.34 s / 2.91 GB | 0.60 s / 0.64 GB |

The advantage grows with the number of columns, because dense cost scales with
`n x p` while sparse scales with the number of non-zeros.

Note the instrument: `gc()` reports R's heap, and Armadillo allocates outside
it, so `gc()`-based memory figures for this package measure everything except
the part that matters.

## Build X sparse, never dense-then-convert

The point is to avoid ever materialising the dense matrix:

```{r, eval = FALSE}
library(Matrix)

X <- sparse.model.matrix(
  ~ pair_type * decade + edu * decade + age_diff * decade + x1,
  data = d
)[, -1, drop = FALSE]          # drop the intercept, not the reference levels

fit <- fastclogit(X, d$choice, d$couple_id,
                  offset  = d$correction,
                  cluster = d$ego_id)
```

`fclogit()`'s formula interface builds `X` **dense**. At this scale, use
`sparse.model.matrix()` and call `fastclogit()` directly.

Dropping `[, -1]` removes the intercept column. Do not instead drop the
reference levels: keeping every level of a factor makes its dummies sum to 1 in
every row, which is a direction constant within every stratum and therefore not
identified. See "Identification" below.

## Sizing a job

Three limits are worth knowing before you queue a four-hour run.

**Non-zeros must stay under 2^31.** The CSR view indexes with `int`.
`ARMA_64BIT_WORD` raises the *virtual cell* ceiling, not this one. Both the R
and C++ sides check it and will tell you the actual count. At 37M rows and ~6
non-zeros per row you are at ~220M, comfortably inside.

**Peak memory is dominated by X.** Sparse storage is roughly
`nnz x 12 bytes` plus the CSR copy. The `p x p` Hessian is negligible until `p`
runs into the thousands.

**Iterations, not rows, drive wall time.** A well-conditioned fit takes 5 to 8
Newton iterations regardless of scale. If yours is taking 40, read
`vignette("convergence-and-diagnostics")` before adding cores.

## Identification at scale

Two failure modes that only bite on large factor-heavy designs.

**Stratum-constant columns.** A covariate constant within every choice set
cancels in the conditional likelihood. `survival::clogit` returns `NA`;
this package returns a ridge-determined value with a meaningless SE, and
`$vcov_singular` does not flag it. In a one-sided choice model this is every
chooser-level main effect. Interact them with an alternative-level covariate,
or drop them.

**Rare cells.** A dummy with a handful of non-zeros in tens of millions of rows
is estimable but fragile. Until v0.5.0 the collinearity screen could delete such
a column outright: it ran QR on a 50,000-row subsample, and a dummy whose 1s all
fell outside it looked like a zero column. At 200,000 rows a dummy with four
1s was dropped in 40% of runs. The screen now forces low-support columns into
the subsample. Still: check `fit$dropped_terms` on every production fit.

## Restricted and offline servers

Where R packages cannot be installed, use the source-able bundle in `mona/`:

```{r, eval = FALSE}
source("load_fastclogit.R")
```

That compiles the four `.cpp` kernels with `Rcpp::sourceCpp()` and sources the
R files. The bundled copies inline the CSR helper rather than `#include`-ing a header,
because R's `make` pipeline treats a `$` in a UNC path (such as
`//server/share/PROJECT$/`) as a make-variable reference and expands it to
nothing, so a header on the include path becomes silently unreachable.

The loader resolves its own directory by checking candidates until one
demonstrably holds `clogit_newton.cpp`, rather than trusting
`sys.frame(1)$ofile`, which names the *outermost* frame, so under nested
`source()` it points at whichever file started the chain. Set
`.this_script_dir` or `CODE_PATH` before sourcing to pin it.

`mona/` is **generated** from `src/` by `tools/make_mona_bundle.R`. Edit `src/`
and regenerate; `tests/testthat/test-mona-bundle.R` fails if the committed
bundle is stale.

## Checklist before a long run

1. Build `X` sparse; never `as.matrix()` it.
2. Confirm `nnz < 2^31`.
3. Drop stratum-constant columns yourself.
4. Fit a subsample first and read `$iter_log`. A specification that needs 40
   iterations at 1% of the data will need them at 100%.
5. After the run: check `$convergence_criterion`, `$dropped_terms` and
   `$vcov_singular` before quoting any number.
